Standartox: A tool to derive exposure endpoints from chemical test results

### 2019-11-13 article/parts/02_methods.tex

\subsection*{Data acquisition and preparation pipe line}
The US Environmental Protection Agency (EPA) releases the EPA ECOTOX data base on a quarterly basis. The latest version, released on 12.09.2019 contains data on 11,756 chemicals, 12,906 taxa, 49,153 references and 952,634 results \citep{usepa_ecotox_2019} constituting a vast collection of ecotoxicological tests, which can be downloaded as pipe delimited ASCII files. Standartox downloads the EPA ECOTOX data base whenever a new version is published and builds into a local PostgreSQL data base \citep{szocs_build_2019}. Subsequently SQL functions for further processing are added and known errors are corrected (and also reported). Lookup tables <PUT LOOKUPTABLES REFERNCE HERE> for filtering and converting the data together with a meta table, containing inter alia the release version of the EPA ECOTOX data base are created as well. In a next step Chemical Abstracts Service (CAS) identifiers are used to query identifier data bases, including the Chemical Identifier Resolver (CIR) service \citep{nationalinstitutesofhealthnih_chemical_2019} and the Pubchem data base \citep{kim_pubchem_2016}. Chemical identifiers (CAS, InchI) and taxonomic names are then used to query chemical and taxonomic specific data bases (Table \ref{tab:data-base-additional}) to add information, such as chemical grouping or organism habitat preferences and global occurrence patterns to the local build. From here on the final Standartox data set is compiled, whereupon concentration and duration units are converted (see tables \ref{tab:conv-concentration}, \ref{tab:conv-duration}) in the process of harmonizing the data. Out of the unique 1229 concentration units in the EPA ECOTOX data base, Standartox retains only those (n = X) which are convertible to one of the following units: ug/l, ul/l, ppb, g/m2, mg/kg, \% and flags the remaining as 'other', resulting in 873473 out of 952625 test results. Likewise, out of the 125 test duration units in the EPA ECOTOX, Standartox retains only those (n = X) which are accurately convertible to hours, leading to 905110 out of 952625 test results). Thereby concentration unit additions such as 'food', soil, ai etc. are neglected since they are also coded in other variables and only hinder the processing of units. Units such as per day (mg/kg/day) are generally excluded. Along with the created Standartox data set, a catalog of it, listing all possible entries and value ranges, for continuous variables is created. As a last step, to guarantee the accuracy of the data, check scripts are run at the end. Single processing scripts are listed in figure \ref{fig:pipeline-tree}. The final Standartox table together with the catalog are exported. Both the APP and the API access it.

@@

\subsection*{Web application (APP) and Application Programming Interface (API)}
The web application and the API load the compressed serialized Standartox data into memory and allow a client to interact with it. The client can then call the functions stx\_filter() and stx\_aggregate() to filter and aggregate the data according to specific parameters (Table \ref{tab:app-parameters}).

\subsubsection*{Web application (APP)}
The interactive APP was built in R, using the shiny web application framework \citep{chang_shiny_2018} which runs with the help of a shiny server \citep{HOW-TO-CITE-SHINY-SERVER} on www.standartox.uni-landau.de. It is set of R functions that allow a user to filter and aggregate the data. The data can then be downloaded.

>>>>

\subsection*{Data acquisition and preparation pipe line}
The US Environmental Protection Agency (EPA) releases the EPA ECOTOX data base on a quarterly basis. The latest version, released on 12.09.2019 contains data on 11,756 chemicals, 12,906 taxa, 49,153 references and 952,634 results \citep{usepa_ecotox_2019} constituting a vast collection of ecotoxicological tests, which can be downloaded as pipe delimited ASCII files. Standartox downloads the EPA ECOTOX data base whenever a new version is published and builds into a local PostgreSQL data base \citep{szocs_build_2019}. Subsequently SQL functions for further processing are added and known errors are corrected (and also reported). Lookup tables <PUT LOOKUPTABLES REFERNCE HERE> for filtering and converting the data together with a meta table, containing inter alia the release version of the EPA ECOTOX data base are created as well. In a next step Chemical Abstracts Service (CAS) identifiers are used to query identifier data bases, including the Chemical Identifier Resolver (CIR) service \citep{nationalinstitutesofhealthnih_chemical_2019} and the Pubchem data base \citep{kim_pubchem_2016}. Chemical identifiers (CAS, InchI) and taxonomic names are then used to query chemical and taxonomic specific data bases (Table \ref{tab:data-base-additional}) to add information, such as chemical grouping or organism habitat preferences and global occurrence patterns to the local build. From here on the final Standartox data set is compiled, whereupon concentration and duration units are converted (see tables \ref{tab:conv-concentration}, \ref{tab:conv-duration}) in the process of harmonizing the data. Out of the unique 1229 concentration units in the EPA ECOTOX data base, Standartox retains only those (n = X) which are convertible to one of the following units: ug/l, g/m2, ppb, mg/kg, \% and flags the remaining as \textit{other}, resulting in 873473 out of 952625 test results. Likewise, out of the 125 test duration units in the EPA ECOTOX, Standartox retains only those (n = X) which are accurately convertible to hours, leading to 905110 out of 952625 test results). Thereby concentration unit additions such as \textit{food}, \textit{soil}, \textit{ai} etc. are neglected since they are also coded in other variables and only hinder the processing of units. Units such as per day (mg/kg/day) are generally excluded. Since in the current version the EPA ECOTOX data base lists 298 distinct entries, endpoints were refined to three endpoint groups NOEX, LOEX and XX50. The former two summarise no-observed-adverse-effect level and lowest-observed-adverse-effect levels, respectively. The latter includes various half maximal effective concentration (e.g. EC50, LC50, LD50 etc.) commonly used in ecotoxicology (Table \ref{tab:???????}). Other endpoints are not included in Standartox. Beyond that, users can filter the data according to the CAS number, the concentration type (e.g. active ingredient, formulation etc.), the chemical class (e.g. fungicides, metals etc.), taxa, organism habitat and region, test duration as well as effect (e.g. Mortality, Growth etc.) groups (See Table \ref{tab:app-parameters}) in one of the two front-ends. Along with the created Standartox data set, a catalog of it, listing all possible entries and value ranges, for continuous variables is created. As a last step, to guarantee the accuracy of the data, check scripts are run at the end. Single processing scripts are listed in figure \ref{fig:pipeline-tree}. The final Standartox table together with the catalog are exported. Both the APP and the API access it.

TODO: How does this fit?
\begin{table}
    \input{article/tables/data_refinement_counts.tex}
    \caption{Reduction of data in the compilation process for Standartox}
    \label{tab:data-refinement}
\end{table}

@@

\subsection*{Web application (APP) and Application Programming Interface (API)}
The web application (APP) and the Application Programming Interface (API) load the compressed serialized Standartox data into memory and allow a client to interact with it. The client can then call the functions stx\_filter() and stx\_aggregate() which filter and aggregate the data according to specific parameters (Table \ref{tab:app-parameters}).

\subsubsection*{Web application (APP)}
The interactive APP was built in R, using the shiny web application framework which runs with the help of a shiny server \citep{chang_shiny_2018} on \url{www.standartox.uni-landau.de}. It is set of R functions that allow a user to filter and aggregate the data. The data can then be downloaded.
